### 2018-01-30 includes/experiment.tex
editing first pass

\section{Experimental Data}
\label{sec:data}

    We generated several problem suites with \fuzzer{} that made one solver perform poorly, but not others. These suites are \theSuites{}. Figure~\ref{fig:cvc-hard} shows the suites that were uniquely difficult for \cvc{}. Figure~\ref{fig:z3str3-hard} shows the suites that were uniquely difficult for \us{}. All experiments were run in series on the same Linux computer, with a timeout of 15 seconds.

    \begin{figure}[H]
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-extracts-small.png}
            \label{fig:concats-extracts-small}
            \caption{Performance on concats-extracts-small}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/different-prefix.png}
            \label{fig:different-prefix}
            \caption{Performance on different-prefix}
        \end{subfigure}
        \caption{Problems hard for \cvc{}}
        \label{fig:cvc-hard}
    \end{figure}

    \begin{figure}[H]
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-balanced.png}
            \label{fig:concats-balanced}
            \caption{Performance on concats-balanced}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-small.png}
            \label{fig:concats-small}
            \caption{Performance on concats-small}
        \end{subfigure}
        \caption{Problems hard for \us{}}
        \label{fig:z3str3-hard}
    \end{figure}

\section{Analyses}
\label{sec:analysis}

    In this section, we analyse our experimental results. We discuss the causes of \us{}'s poor performance on the two problem suites: \textit{concats-balanced} and \textit{concats-big}. We were not equipped well enough to debug the behaviours of the other solvers, but we provide our data in hopes that they will be useful to the solvers' authors.

    \subsection{\us{} on concats-balanced}

        \todo{Dmitry with help: Explain why \us{} performs poorly on this suite.}

>>>>

\section{Experimental Data}
\label{sec:data}
\vspace{-0.1in}

    We generated several problem suites with \fuzzer{} that made one solver perform poorly, but not others. These suites are \theSuites{}. Figure~\ref{fig:cvc-hard} shows the suites that were uniquely difficult for \cvc{}. Figure~\ref{fig:z3str3-hard} shows the suites that were uniquely difficult for \us{}. All experiments were run in series on the same Linux computer, with a timeout of 15 seconds. \todo{Specify CPU, memory}

    \begin{figure}[h]
        \vspace{-0.2in}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-extracts-small.png}                   
            \vspace{-0.25in}
            \caption{Performance on concats-extracts-small} 
            \label{fig:concats-extracts-small}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/different-prefix.png}            
            \vspace{-0.25in}
            \caption{Performance on different-prefix}
            \label{fig:different-prefix}
        \end{subfigure}
        \vspace{-0.1in}
        \caption{Problems hard for \cvc{}}
        \label{fig:cvc-hard}
        \vspace{-0.3in}
    \end{figure}

    \begin{figure}[h]
    \vspace{-0.25in}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-balanced.png}
            \label{fig:concats-balanced}
            \vspace{-0.25in}
            \caption{Performance on concats-balanced}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-small.png}
            \label{fig:concats-small}
            \vspace{-0.25in}
            \caption{Performance on concats-small}
        \end{subfigure}
        \vspace{-0.1in}
        \caption{Problems hard for \us{}}
        \label{fig:z3str3-hard}
        \vspace{-0.3in}
    \end{figure}

\section{Usefulness to \us{}: A Case Study}
\vspace{-0.1in}

        We describe the defects and missing heuristics exposed by \fuzzer{} to illustrate why \fuzzer{} is a valuable asset in solver development.

\vspace{-0.15in}
\subsection{Defects discovered and fixed in \us{}}
\vspace{-0.1in}

The first issue we found caused \us{} to perform poorly when equivalence classes 
of arithmetic terms were very large (e.g. the lengths of many strings were equal
to 0). This issue was detected by the \texttt{Concats} problem suite. The slowdown 
was caused by a loop over terms in the same equivalence class to search for an 
integer constant. The fix was to check the root of the equivalence class for the 
integer constant, as Z3 makes ``interpreted terms'' (e.g. constants) the root of 
equivalence classes whenever possible. This issue was last observed in commit 
\texttt{6308636}, and was fixed in commit \texttt{3865c45}.

The second issue, which was exposed by the \texttt{Lengths} suite, caused \us{} to 
perform poorly when a variable has a large length constraint.
The slowdown was caused by the solver performing a search for a satisfying length
even when the input formula explicitly constrained the exact length of a string term.
The fix was to check whether the integer solver had already assigned a length value
before searching for length, and skip the search if this was the case.
This issue was last observed in commit \texttt{66bc68f}, and was fixed in commit 
\texttt{7b536e9}.    
%The slowdown was  caused by the solver performing a linear search for a satisfying length when 
%producing a model for the variable. This issue was an inconsistency between the 
%implementation and the behavior described in the \us{} paper \cite{z3str3}. 
%This issue was last observed in commit \texttt{66bc68f}, and was fixed in commit 
%\texttt{7b536e9}.

\vspace{-0.15in}    
\subsection{Solving heuristics}
\vspace{-0.1in}

In order to improve the efficiency of any solver, discovering new heuristics is crucial.
As an illustrative example, consider detection of conflicting constraints.
A solver that follows the string constraint solving algorithm and gradually propagates
the (dis)equalities discovered along the way will eventually detect a conflict.
However, in many scenarios, a conflict can be discovered ``early'' before trying to
figure out the relationships among subterms and propagating the facts.
The benefits of this eager approach are clear: this saves the effort of reasoning about
subterms and propagating observations, as a conflict will eventually be reached in either case,
forcing conflict analysis and a backjump.
As reported in existing works, heuristics of this form usually lead to significant improvements.
However, the process of discovering such heuristics is extremely challenging,
as they are usually learned from a class of examples whose commonalities
are not always obvious or immediately indicative of such inefficiencies.

The instances in the \texttt{concats-big} suite generated by \fuzzer{} helped us 
discover a missing heuristic in \us{}. In particular, \us{} didn't make full use 
of the solving context (e.g. some terms are empty strings) to simplify the 
concatenations of a long list of string terms before trying to reason about the 
equivalences among subterms. \us{} therefore introduced a large number of unnecessary
intermediate variables and propagations.
As an immediate consequence of the \texttt{concats-big} benchmarks, we were able to identify
the factors common to these instances and rapidly formulate a hypothesis as to why they were
performing poorly. This allowed us to move forward with the development of new heuristics
to resolve these issues and achieve better performance.

% In this section, we analyse our experimental results. We discuss the causes of 
% \us{}'s poor performance on the two problem suites: \textit{concats-balanced} 
% and \textit{concats-big}. We were not equipped well enough to debug the 
% behaviours of the other solvers, but we provide our data in hopes that they will 
% be useful to the solvers' authors.

%     \subsection{\us{} on concats-balanced}
%         \todo{Dmitry with help: Explain why \us{} performs poorly on this suite.}

### 2018-01-30 includes/introduction.tex
editing first pass

    Many SMT solvers supporting the theory of strings (e.g., CVC4 \cite{cvc4},  Z3str2 \cite{z3str2}, Norn \cite{norn} and Z3str3 \cite{z3str3}) have been developed over the past decade. To validate and benchmark their solvers, many authors have relied on hand-crafted input suites \cite{cvc4-tests} \cite{z3str3-tests} \cite{z3str2-tests} or real-world examples from industrial applications \cite{kaluza} \cite{kausler}. These test suites have helped developers identify implementation defects and develop more sophistic solving heuristics. Although extremely helpful, these suites can only cover a limited range of the problem space because they are either manually created or translated from carefully developed programs.

    Fuzzing, on the other hand, has already been widely used in similar scenarios to systematically generate more diverse test cases. Guided by different rules, it can also be applied to generate different instances that are particularly designed to challenge selected capacities. Inspired by traditional fuzzers, we introduce a modular fuzzer for string solvers, \fuzzer{}, and describe its utility as an exploratory validation tool. We demonstrate its effectiveness by presenting defects and limitations we were able to identify in leading solvers.
%     Moreover, a standard input language for problem instances, \smtfull{}\cite{smt}, has been developed and widely adopted.

@@

        \begin{enumerate}
            \item \textbf{The \fuzzer{} tool}: We describe a modular fuzzer which can transform and generate \smtfull{} string and regex problems. We briefly document its components and modular architecture. We provide example usages to demonstrate its usefulness as an exploratory tool for validating solvers. We additionally detail its regex-generating and transforming capabilities.
            \item \textbf{A repository of \smtfull{} problems}: We present a repository of \smtfull{} string and regex problem suites we generated using \fuzzer{}. We demonstrate the usefulness of these suites by revealing specific defects they helped uncover and fix in \us{}.
            \item \textbf{Experimental Results}: We compare \theSolvers{} on the \fuzzer{} suites and selectively present the experimental result on \theSuites{}. We highlighted these suites because they uniquely made some solvers perform poorly, but not others.
            \item \textbf{Analysis}: We analyze our experimental results by inspecting the execution traces of the solvers on the given problems. For \us{}, we pinpoint algorithmic limitations that cause the poor performance. For other solvers, we draw superficial conclusions based solely on the execution traces, and invite their authors to conduct further analysis.
        \end{enumerate}

>>>>

    Many SMT solvers supporting the theory of strings (e.g., CVC4 \cite{cvc4},  Z3str2 \cite{z3str2}, Norn \cite{norn} and Z3str3 \cite{z3str3}) have been developed over the past decade. To validate and benchmark their solvers, many authors have relied on hand-crafted input suites~\cite{cvc4-tests,z3str3-tests,z3str2-tests} or real-world examples from industrial applications~\cite{kaluza,kausler}. These test suites have helped developers identify implementation defects and develop more sophisticated solving heuristics. Although extremely helpful, these suites can only cover a limited range of the problem space because they are either manually created or translated from carefully developed programs.

    Fuzzing, on the other hand, has already been widely used in similar scenarios to systematically generate more diverse test cases. Guided by different rules, it can also be applied to generate different instances that are particularly designed to challenge selected parts of the solvers. Inspired by traditional fuzzers, we introduce a modular fuzzer for string solvers, \fuzzer{}, and describe its utility as an exploratory validation tool. We demonstrate its effectiveness by presenting defects and limitations we were able to identify in leading solvers.
%     Moreover, a standard input language for problem instances, \smtfull{}\cite{smt}, has been developed and widely adopted.

@@

        \begin{enumerate}
            \item \textbf{The \fuzzer{} tool}: We describe a modular fuzzer which can transform and generate \smtfull{} string and regex problems. We briefly document its components and modular architecture. We provide example use cases to demonstrate its usefulness as an exploratory tool for validating solvers. We additionally detail its regex generation and transformation capabilities.
            \item \textbf{A repository of \smtfull{} problems}: We present a repository of \smtfull{} string and regex problem suites we generated using \fuzzer{}. We demonstrate the usefulness of these suites by revealing specific defects they helped uncover and fix in \us{}.
            \item \textbf{Experimental Results}: We compare \theSolvers{} on the \fuzzer{} suites and selectively present the experimental result on \theSuites{}. We highlighted these suites because they uniquely made some solvers perform poorly, but not others.
            \item \textbf{Analysis}: We analyze our experimental results by inspecting the execution traces of the solvers on the given problems. For \us{}, we pinpoint algorithmic limitations that cause poor performance. For other solvers, we draw superficial conclusions based solely on the execution traces, and invite their authors to conduct further analysis.
        \end{enumerate}

### 2018-01-30 includes/prior.tex
editing first pass

    Naturally, many solver developers author their own test suites to validate their solvers \cite{cvc4-tests} \cite{z3str3-tests} \cite{z3str2-tests}. In addition, several popular problem suites are publicly available for solver validation, such as the Kaluza \cite{kaluza} and Kausler \cite{kausler} suites.

    There are likewise several fuzzers and problem generators currently available, but none of them can generate or transform string and regex problems. For example the FuzzSMT\cite{fuzzsmt} tool generates \smt{} problems with bit vectors and arrays, but does not support strings or regexes. The SMTpp\cite{smtpp} tool preprocesses and simplifies problems, but does not generate new ones or fuzz existing ones.

>>>>

    Naturally, many solver developers author their own test suites to validate their solvers~\cite{cvc4-tests,z3str3-tests,z3str2-tests}. In addition, several popular problem suites are publicly available for solver validation, such as the Kaluza~\cite{kaluza} and Kausler~\cite{kausler} suites.

    There are likewise several fuzzers and problem generators currently available, but none of them can generate or transform string and regex problems. For example, the FuzzSMT~\cite{fuzzsmt} tool generates \smt{} problems with bit-vectors and arrays, but does not support strings or regexes. The SMTpp~\cite{smtpp} tool pre-processes and simplifies problems, but does not generate new ones or fuzz existing ones.

### 2018-01-30 includes/problems.tex
editing first pass

\subsection{Generated suite}
\vspace{-0.1in}
    We generated several suites of string and regex problems using \fuzzer{}\footnote{For review anonymity, we removed the link to the source code. It will be provided separately in the artifact submission.}. Table~\ref{tbl:generated} describes the suite \texttt{generated} by \generator{}.

>>>>

In this section, we describe the benchmark suites that we generated with \fuzzer{} on which the experimental evaluation was conducted.

\subsection{Generated suite}
\vspace{-0.1in}
    We generated several suites of string and regex problems using \fuzzer{}\footnote{For review anonymity, we removed the link to the source code and the generated suites. They will be provided separately in the artifact submission.}. Table~\ref{tbl:generated} describes the suite \texttt{generated} by \generator{}.

\todo{Specify how many total instances were generated.}

### 2018-01-30 includes/stringfuzz.tex
editing first pass

    Below are descriptions of the tools that constitute \fuzzer{}:

\vspace{-0.15in}
\subsection{\generator{}}
% \vspace{-0.1in}

\generator{} generates \smt{} problems. It supports several generators and options that specify its output. Details can be found in Table~\ref{tbl:generators}.

@@

\vspace{-0.15in}
\subsection{\transformer{}}
% \vspace{-0.1in}

@@

\vspace{-0.15in}
\subsection{\texttt{stringstats}}

@@

    \section{Regex Fuzzing Capabilities}
    \vspace{-0.1in}
        \fuzzer{} can generate and transform problems with regular expression constraints. The command \texttt{stringfuzzg regex} invokes the regex problem generator. It generates a problem of the form: %\footnote{Variable declarations omitted for brevity.}:
        \begingroup
        \footnotesize
        \begin{align*}
            & \texttt{(assert (str.in.re X}\; R_0\; \texttt{))} \\
            & \texttt{(assert (str.in.re X}\; R_n\; \texttt{))}* \\
            & \texttt{(assert (<= Min (str.len X)))}? \\
            & \texttt{(assert (<= (str.len X)) Max)}?
        \end{align*}
        \endgroup
        where $R_i \in RegEx$, and $Min, Max \in Int$. More simply, the problem is a set of one or more regex constraints on a single variable, with optional maximum and minimum lengths constraints. The regex constraints $R$ are each of the form:
        \begingroup
        \footnotesize
        \begin{align*}
            & \texttt{(re.++}\; \ T_0\; \  T_1\; \texttt{...}\; T_n\; \texttt{)}
        \end{align*}
        \endgroup
        and each $T_i$ is a recursive term of the form:
        \begingroup
        \footnotesize
        \begin{align*}
            & \texttt{(re.*}\; T_{i_j}\; \texttt{) | (re.+}\; T_{i_j}\; \texttt{) | (re.union}\; T_{i_{j_1}}\; T_{i_{j_2}}\; \texttt{)}
        \end{align*}
        \endgroup
        where $j$ a specified depth of recursion. Terms at depth 0 are regex constants. Informally, this form describes a concatenation of regex terms. Each term is a random nested regex operator (chosen from regex Kleene star, repetition, and union), up to a specified depth, terminating in a regex literal. Below is example regex (with spacing added for clarity) of depth 2 produced by this scheme:
        \begingroup
        \footnotesize
        \begin{align*}
            & ((\texttt{a}|\texttt{b})|(\texttt{cc})+)\quad ((\texttt{ddd})*)+\quad ((\texttt{ee})+|(\texttt{fff})*)
        \end{align*}
        \endgroup

        {\it \fuzzer{} can also transform the instances it generates}. All of the transformers in \transformer{} can operate on regex terms. They can multiply and permute regex literals (e.g. \texttt{reverse}, \texttt{multiply}, \texttt{translate}), randomly replace regex operators (e.g. \texttt{fuzz}), and manipulate the syntax tree (e.g. \texttt{graft}, \texttt{rotate}).

\section{Example Usages}
\vspace{-0.1in}

        \fuzzer{} can be used to generate problems to be saved as static suites. It can also be used in a randomised fashion for exploratory debugging. As an example, the following usage can feed random problems to \cvc{} until it returns an error:
\begin{center}
\vspace{-0.05in}
\begin{varwidth}{\linewidth}
{\scriptsize\begin{verbatim}while stringfuzzg -r random | tee problem.smt25 | cvc4 --lang smt2.5 --tlimit=5000; do
    sleep 0
done
\end{verbatim}}
\end{varwidth}
\end{center}
        Some \unix{} scripting can yield other interesting results. For example, the following script finds a regex problem on which \us{} times out after 5 seconds:
\begin{center}
\vspace{-0.1in}
\begin{varwidth}{\linewidth}
        {\scriptsize\begin{verbatim}while true; do
    stringfuzzg -r regex -c -a increasing -r 2 -t 5 -d 3 > problem.smt25
    result=`z3str3 -T:5 problem.smt25`
    echo $result
    if [ "$result" == "timeout" ]; then
        cat hard-problem.smt2
        break
    fi
done\end{verbatim}}
\end{varwidth}
\end{center}

>>>>

    \fuzzer{} is composed of the following tools:

\vspace{-0.15in}
\subsubsection{\generator{}:}
% \vspace{-0.1in}

This tool generates \smt{} problems. It supports several generators and options that specify its output. Details can be found in Table~\ref{tbl:generators}.

@@

\vspace{-0.15in}
\subsubsection{\transformer{}:}
% \vspace{-0.1in}

@@

\vspace{-0.15in}
\subsubsection{\texttt{stringstats}:}

@@

    \section{Regex Fuzzing Capabilities}
    \vspace{-0.1in}
        \fuzzer{} can generate and transform problems with regular expression constraints. The command \texttt{stringfuzzg regex} invokes the regex problem generator. It generates a problem of the form: %\footnote{Variable declarations omitted for brevity.}:
        \begingroup
        \footnotesize
        \begin{align*}
            & \texttt{(assert (str.in.re X}\; R_0\; \texttt{))} \\
            & \texttt{(assert (str.in.re X}\; R_n\; \texttt{))}* \\
            & \texttt{(assert (<= Min (str.len X)))}? \\
            & \texttt{(assert (<= (str.len X)) Max)}?
        \end{align*}
        \endgroup
        where $R_i \in RegEx$, and $Min, Max \in Int$. More simply, the problem is a set of one or more regex constraints on a single variable, with optional maximum and minimum length constraints. The regex constraints $R$ are each of the form:
        \begingroup
        \footnotesize
        \begin{align*}
            & \texttt{(re.++}\; \ T_0\; \  T_1\; \texttt{...}\; T_n\; \texttt{)}
        \end{align*}
        \endgroup
        and each $T_i$ is a recursive term of the form:
        \begingroup
        \footnotesize
        \begin{align*}
            & \texttt{(re.*}\; T_{i_j}\; \texttt{) | (re.+}\; T_{i_j}\; \texttt{) | (re.union}\; T_{i_{j_1}}\; T_{i_{j_2}}\; \texttt{)}
        \end{align*}
        \endgroup
        where $j$  is the specified depth of recursion. Terms at depth 0 are regex constants. Informally, this form describes a concatenation of regex terms. Each term is a random nested regex operator (chosen from regex Kleene star, repetition, and union), up to a specified depth, terminating in a regex literal. Below are several example regexes (with spacing added for clarity) of depth 2 produced by this scheme:
        \begingroup
        \footnotesize
        \begin{align*}
            & ((\texttt{a}|\texttt{b})|(\texttt{cc})+)\quad ((\texttt{ddd})*)+\quad ((\texttt{ee})+|(\texttt{fff})*)
        \end{align*}
        \endgroup

        \fuzzer{} can also transform the instances it generates. All of the transformers in \transformer{} can operate on regex terms. They can multiply and permute regex literals (e.g. \texttt{reverse}, \texttt{multiply}, \texttt{translate}), randomly replace regex operators (e.g. \texttt{fuzz}), and manipulate the syntax tree (e.g. \texttt{graft}, \texttt{rotate}).

\section{Example Use Cases}
\vspace{-0.1in}

        \fuzzer{} can be used to generate problems to be saved as static suites. It can also be used in a randomised fashion for exploratory debugging. As an example, \fuzzer{} can be invoked as follows to feed random problems to \cvc{} until it returns an error:
\begin{center}
\vspace{-0.05in}
\begin{varwidth}{\linewidth}
{\scriptsize\begin{verbatim}while stringfuzzg -r random | tee problem.smt25 | cvc4 --lang smt2.5 --tlimit=5000; do
    sleep 0
done
\end{verbatim}}
\end{varwidth}
\end{center}
        Some \unix{} scripting can yield other interesting results. For example, the following script finds a regex problem on which \us{} times out after 5 seconds:
\begin{center}
\vspace{-0.1in}
\begin{varwidth}{\linewidth}
        {\scriptsize\begin{verbatim}while true; do
    stringfuzzg -r regex -c -a increasing -r 2 -t 5 -d 3 > problem.smt25
    result=`z3str3 -T:5 problem.smt25`
    echo $result
    if [ "$result" == "timeout" ]; then
        cat problem.smt25
        break
    fi
done\end{verbatim}}
\end{varwidth}
\end{center}
